---
title: "Analysis of Police Violence in the US"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggplot2)
library(readr)
library(readxl)
library(patchwork)
library(leaflet)
library(rvest)
#you'll need to install the packages below in order to run animated plots or knit
library(gganimate)
library(gifski)
library(png)
library(plotly)

knitr::opts_chunk$set(
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_color_viridis_d
scale_fill_discrete = scale_fill_viridis_d

mpv_df =
  read_csv("./data/mpv_final.csv")
```

### Total Number of People Killed by Police by Department
```{r, warning = FALSE, message = FALSE}
mpv_df %>% 
  count(police_dept) %>% 
  rename(total_killed_pd = n) %>% 
  filter(total_killed_pd >= 49) %>% 
  mutate(
    police_dept = factor(police_dept),
    police_dept = fct_reorder(police_dept, total_killed_pd)
  ) %>% 
   plot_ly(
      y = ~total_killed_pd, x = ~police_dept, color = ~police_dept,
      type = "bar", colors = "viridis")
```
This plot depicts the total number of people killed from 2013 through 2020 broken down by police department; only departments with a count higher than 49 are included to show the highest offenders. Based on this graph, we can see that Los Angeles has the highest count, especially once the LA County and LA city departments are combined together for a total that is double the city with the second highest count.   


The plots below visualize police killings by state. One includes all states with data, the other includes the top 10 states. 

```{r}
state_plot = 
  mpv_df %>% 
  count(state) %>% 
  rename(total_killed_st = n) %>% 
  arrange(desc(total_killed_st)) %>% 
  mutate(
    state = factor(state),
    state = fct_reorder(state, total_killed_st)
  ) %>% 
  ggplot(aes(x = state, y = total_killed_st, fill = state)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

state_subset_plot = 
  mpv_df %>% 
  count(state) %>% 
  rename(total_killed_st = n) %>% 
  arrange(desc(total_killed_st)) %>% 
  filter(total_killed_st >= 230) %>% 
  mutate(
    state = factor(state),
    state = fct_reorder(state, total_killed_st)
  ) %>% 
  ggplot(aes(x = state, y = total_killed_st, fill = state)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

state_plot + state_subset_plot
```

The plot below shows police killings over time until 2019 (I excluded 2020, because otherwise it looks like 2020 has a drop in police killings).

```{r}
year_plot = 
  mpv_df %>% 
  count(year) %>% 
  rename(total_killed = n) %>% 
  filter(year < 2020) %>% 
  ggplot(aes(x = year, y = total_killed)) + geom_point() +
  geom_smooth(method = lm, se = FALSE) +
  scale_y_continuous(limit = c(0, 1400))

year_plot
```

### Breakdown of charges laid against police

```{r}
mpv_case = 
  mpv_df %>% 
  count(criminal_charges) %>% 
  mutate(
    charges = case_when(
      startsWith(criminal_charges, "Charged") ~ "Yes",
      startsWith(criminal_charges, "No") ~ "No",
    )) %>% 
  ggplot(aes(x = charges, y = n, color = charges, fill = charges)) +
  geom_col() +
  labs(
    x = "Charges Brought Up Against Police" ,
    y = "Total"
  ) +
  theme(legend.title = element_blank()) 

mpv_case
```
The above plot shows the breakdown of whether charges were pressed police after each case. In total, `r mpv_df %>% filter(criminal_charges == "No known charges") %>% count()`charges were brought up against police, compared to the `r mpv_df %>% filter(criminal_charges == "Charged with a crime") %>% count()` cases where police were not charged. 


Scatter plot showing killings over time separated by race

```{r}
anim_scatter =
  mpv_df %>% 
  group_by(year, race) %>% 
  count() %>% 
  ggplot(aes(x = year, y = n, color = race)) +
  geom_line() +
  geom_point() +
  labs(title = "Number of Police Killings Over Time by Race",
       x = "Year",
       y = "Number of Police Killings") +
  transition_reveal(year)

animate(anim_scatter, renderer = gifski_renderer())

# might be worth excluding 2020 because of incomplete data, it makes it look like the shootings are trending down
```


