---
title: "Data Wrangling for All Datasets Used"
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
library(readxl)
library(readr)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_color_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

## MPV Dataset

```{r, warning=FALSE}
mpv_df =
  read_excel("./data/MPVDatasetDownload.xlsx",
             sheet = "2013-2020 Police Killings") %>%
  janitor::clean_names() %>% 
  separate(date_of_incident_month_day_year, c("year", "month", "day")) %>% 
  mutate(
    age = as.numeric(victims_age),
    gender  = victims_gender,
    race = victims_race,
    police_dept = agency_responsible_for_death,
    description = a_brief_description_of_the_circumstances_surrounding_the_death,
    disposition = official_disposition_of_death_justified_or_other,
    year = as.numeric(year),
    month = as.numeric(month),
    day = as.numeric(day)) %>%
  select(
    age, gender, race, police_dept, disposition, year,
    month, day, city, state, county, cause_of_death, criminal_charges,
    symptoms_of_mental_illness) %>% 
  view()
```

In the above code chunk, the MPV dataset has been manipulated and tidied by reading in the excel sheet and cleaning the variable names, separating out the date variable, changing the names of some of the variables and changing some variables to be numeric, and selecting only the variables of importance to our analysis for inclusion in the tidied dataset.

## Lat/Long Dataset

```{r}
lat_long =
  read_csv("./data/uscities.csv") %>% 
  mutate(state = state_id) %>% 
  select(city, lat, lng, state)

mpv_final = left_join(mpv_df, lat_long, by = c("city" = "city", "state" = "state"))

write_csv(mpv_final, "./data/mpv_final.csv")
```


## USA October 2020 Dataset

```{r usa, warning=FALSE}
usa_df =
  read_excel("./data/usa_2020_oct.xlsx") %>%
  janitor::clean_names() %>% 
  separate(event_date, c("year", "month", "day")) %>%
  mutate(
    state = admin1,
    county = admin2,
    city = location,
    latitude = as.numeric(latitude),
    longitude = as.numeric(longitude),
    year = as.numeric(year),
    month = as.numeric(month),
    day = as.numeric(day),
    fatalities = as.numeric(fatalities)) %>%
  select(
    year, month, day, event_type, sub_event_type, state, county, city, latitude, longitude, fatalities)

write_csv(usa_df, "./data/usa_tidy.csv")
```

## ACS Datasets

#### Age and Sex

Using the ACS age/sex dataset to create an age only dataset:
```{r}
# getting the total ages by county and tidying
age_total_df =
  read_csv("./data/acs_age_county.csv", skip = 1) %>% 
  janitor::clean_names() %>% 
  select(c(1:3, starts_with("estimate_total_total"))) %>% 
  select(-c("estimate_total_total_population", contains(c("selected_age_categories", "summary_indicators", "percent_allocated")))) %>% 
  separate(geographic_area_name, c("county", "state"), sep = ",") %>%
  pivot_longer(
    4:21,
    names_to = "age",
    values_to = "total",
    names_prefix = "estimate_total_total_population_age_"
  ) %>%
  mutate(
    age = str_replace_all(age, "_", " "),
    age = str_replace(age, "years", "")
  ) %>% view()

# getting the percentage of ages by county and tidying
age_percent_df =
  read_csv("./data/acs_age_county.csv", skip = 1) %>% 
  janitor::clean_names() %>% 
  select(c(1:3, starts_with("estimate_percent_total"))) %>% 
  select(-c("estimate_total_total_population", "estimate_percent_total_population", contains(c("selected_age_categories", "summary_indicators", "percent_allocated")))) %>% 
  separate(geographic_area_name, c("county", "state"), sep = ",") %>%
  pivot_longer(
    4:21,
    names_to = "age",
    values_to = "percent",
    names_prefix = "estimate_percent_total_population_age_"
  ) %>%
  mutate(
    age = str_replace_all(age, "_", " "),
    age = str_replace(age, "years", "")
  ) %>% view()

# joining the total and percentages together into one dataset
age_df = 
  left_join(age_total_df, age_percent_df, by = c("id", "county", "state", "age")) %>% view()
  
# export tidy age dataset
write_csv(age_df, "./data/age_tidy.csv")
```

Using the ACS age/sex dataset to create an age by sex dataset:
```{r}
# read in total for males by age

male_total_df =
  read_csv("./data/acs_age_county.csv", skip = 1) %>% 
  janitor::clean_names() %>% 
  select(c(1:3, starts_with("estimate_male_total"))) %>% 
  select(-c("estimate_total_total_population", "estimate_male_total_population", contains(c("selected_age_categories", "summary_indicators", "percent_allocated")))) %>% 
  separate(geographic_area_name, c("county", "state"), sep = ",") %>% 
  pivot_longer(
    4:21,
    names_to = "age",
    values_to = "total",
    names_prefix = "estimate_male_total_population_age_"
  ) %>%
  mutate(
    age = str_replace_all(age, "_", " "),
    age = str_replace(age, "years", ""),
    sex = "male"
  ) %>% view()

# read in total for females by age

female_total_df =
  read_csv("./data/acs_age_county.csv", skip = 1) %>% 
  janitor::clean_names() %>% 
  select(c(1:3, starts_with("estimate_female_total"))) %>% 
  select(-c("estimate_total_total_population", "estimate_female_total_population", contains(c("selected_age_categories", "summary_indicators", "percent_allocated")))) %>% 
  separate(geographic_area_name, c("county", "state"), sep = ",") %>%
  pivot_longer(
    4:21,
    names_to = "age",
    values_to = "total",
    names_prefix = "estimate_female_total_population_age_"
  ) %>%
  mutate(
    age = str_replace_all(age, "_", " "),
    age = str_replace(age, "years", ""),
    sex = "female"
  ) %>% view()

# bind male/female total dfs together

sex_total_df =
  bind_rows(male_total_df, female_total_df) %>% 
  relocate(id, county, state, age, sex) %>% 
  view()

# read in percent for males by age

male_percent_df =
  read_csv("./data/acs_age_county.csv", skip = 1) %>% 
  janitor::clean_names() %>% 
  select(c(1:3, starts_with("estimate_percent_male"))) %>% 
  select(-c("estimate_total_total_population", "estimate_percent_male_total_population", contains(c("selected_age_categories", "summary_indicators", "allocated")))) %>% 
  separate(geographic_area_name, c("county", "state"), sep = ",") %>% 
  pivot_longer(
    4:21,
    names_to = "age",
    values_to = "percent",
    names_prefix = "estimate_percent_male_total_population_age_"
  ) %>%
  mutate(
    age = str_replace_all(age, "_", " "),
    age = str_replace(age, "years", ""),
    sex = "male"
  ) %>% view()

# read in percent for females by age

female_percent_df =
  read_csv("./data/acs_age_county.csv", skip = 1) %>% 
  janitor::clean_names() %>% 
  select(c(1:3, starts_with("estimate_percent_female"))) %>% 
  select(-c("estimate_total_total_population", "estimate_percent_female_total_population", contains(c("selected_age_categories", "summary_indicators", "percent_allocated")))) %>% 
  separate(geographic_area_name, c("county", "state"), sep = ",") %>%
  pivot_longer(
    4:21,
    names_to = "age",
    values_to = "percent",
    names_prefix = "estimate_percent_female_total_population_age_"
  ) %>%
  mutate(
    age = str_replace_all(age, "_", " "),
    age = str_replace(age, "years", ""),
    sex = "female"
  ) %>% view()

# bind male/female percent dfs together

sex_percent_df =
  bind_rows(male_percent_df, female_percent_df) %>% 
  relocate(id, county, state, age, sex) %>% 
  view()

# join percent and total dfs together

sex_df = 
  left_join(sex_total_df, sex_percent_df, by = c("id", "county", "state", "age", "sex")) %>% view()
  
# export sex by age dataset
write_csv(sex_df, "./data/age_sex_tidy.csv")
```

#### Age dataset (not tidy for analysis potentially)

```{r, warning = FALSE, message = FALSE}
age_df =
  read_csv("./data/acs_age_county.csv", skip = 1, na = "null") %>% 
  janitor::clean_names() %>% 
  select(-c(starts_with("margin_of_error"))) %>% 
  select(geographic_area_name,
         estimate_total_total_population_age_under_5_years,
         estimate_total_total_population_age_5_to_9_years,
         estimate_total_total_population_age_10_to_14_years,
         estimate_total_total_population_age_15_to_19_years,
         estimate_total_total_population_age_20_to_24_years,
         estimate_total_total_population_age_25_to_29_years,
         estimate_total_total_population_age_30_to_34_years,
         estimate_total_total_population_age_35_to_39_years,
         estimate_total_total_population_age_40_to_44_years,
         estimate_total_total_population_age_45_to_49_years,
         estimate_total_total_population_age_50_to_54_years,
         estimate_total_total_population_age_55_to_59_years,
         estimate_total_total_population_age_60_to_64_years,
         estimate_total_total_population_age_65_to_69_years,
         estimate_total_total_population_age_70_to_74_years,
         estimate_total_total_population_age_75_to_79_years,
         estimate_total_total_population_age_80_to_84_years,
         estimate_total_total_population_age_85_years_and_over,
         estimate_percent_total_population_age_under_5_years,
         estimate_percent_total_population_age_5_to_9_years,
         estimate_percent_total_population_age_10_to_14_years,
         estimate_percent_total_population_age_15_to_19_years,
         estimate_percent_total_population_age_20_to_24_years,
         estimate_percent_total_population_age_25_to_29_years,
         estimate_percent_total_population_age_30_to_34_years,
         estimate_percent_total_population_age_35_to_39_years,
         estimate_percent_total_population_age_40_to_44_years,
         estimate_percent_total_population_age_45_to_49_years,
         estimate_percent_total_population_age_50_to_54_years,
         estimate_percent_total_population_age_55_to_59_years,
         estimate_percent_total_population_age_60_to_64_years,
         estimate_percent_total_population_age_65_to_69_years,
         estimate_percent_total_population_age_70_to_74_years,
         estimate_percent_total_population_age_75_to_79_years,
         estimate_percent_total_population_age_80_to_84_years,
         estimate_percent_total_population_age_85_years_and_over) %>% 
  separate(geographic_area_name, c("county", "state"), sep = ",") %>% 
  rename(age_under_5 = estimate_total_total_population_age_under_5_years,
         age_5_9 = estimate_total_total_population_age_5_to_9_years,
         age_10_14 = estimate_total_total_population_age_10_to_14_years,
         age_15_19 = estimate_total_total_population_age_15_to_19_years,
         age_20_24 = estimate_total_total_population_age_20_to_24_years,
         age_25_29 = estimate_total_total_population_age_25_to_29_years,
         age_30_34 = estimate_total_total_population_age_30_to_34_years,
         age_35_39 = estimate_total_total_population_age_35_to_39_years,
         age_40_44 = estimate_total_total_population_age_40_to_44_years,
         age_45_49 = estimate_total_total_population_age_45_to_49_years,
         age_50_54 = estimate_total_total_population_age_50_to_54_years,
         age_55_59 = estimate_total_total_population_age_55_to_59_years,
         age_60_64 = estimate_total_total_population_age_60_to_64_years,
         age_65_69 = estimate_total_total_population_age_65_to_69_years,
         age_70_74 = estimate_total_total_population_age_70_to_74_years,
         age_75_79 = estimate_total_total_population_age_75_to_79_years,
         age_80_84 = estimate_total_total_population_age_80_to_84_years,
         age_85_over = estimate_total_total_population_age_85_years_and_over,
         age_pct_under_5 = estimate_percent_total_population_age_under_5_years,
         age_pct_5_9 = estimate_percent_total_population_age_5_to_9_years,
         age_pct_10_14 = estimate_percent_total_population_age_10_to_14_years,
         age_pct_15_19 = estimate_percent_total_population_age_15_to_19_years,
         age_pct_20_24 = estimate_percent_total_population_age_20_to_24_years,
         age_pct_25_29 = estimate_percent_total_population_age_25_to_29_years,
         age_pct_30_34 = estimate_percent_total_population_age_30_to_34_years,
         age_pct_35_39 = estimate_percent_total_population_age_35_to_39_years,
         age_pct_40_44 = estimate_percent_total_population_age_40_to_44_years,
         age_pct_45_49 = estimate_percent_total_population_age_45_to_49_years,
         age_pct_50_54 = estimate_percent_total_population_age_50_to_54_years,
         age_pct_55_59 = estimate_percent_total_population_age_55_to_59_years,
         age_pct_60_64 = estimate_percent_total_population_age_60_to_64_years,
         age_pct_65_69 = estimate_percent_total_population_age_65_to_69_years,
         age_pct_70_74 = estimate_percent_total_population_age_70_to_74_years,
         age_pct_75_79 = estimate_percent_total_population_age_75_to_79_years,
         age_pct_80_84 = estimate_percent_total_population_age_80_to_84_years,
         age_pct_85_over = estimate_percent_total_population_age_85_years_and_over
  )

```

#### ACS Income

```{r}
income_df =
  read_csv("./data/acs_income_county.csv", skip = 1) %>% 
  janitor::clean_names() %>% 
  separate(geographic_area_name, c("county", "state"), sep = ",") %>% 
  select(
      -c(id, 
         starts_with("margin_of_error"), 
         starts_with("estimate_families"), 
         starts_with("estimate_married"), 
         starts_with("estimate_nonfamily"), 
         starts_with("estimate_households_percent"))
      ) %>%
  mutate(
    disparity_value = estimate_households_mean_income_dollars - estimate_households_median_income_dollars) 

# mutate(
#    income_class =
#     case_when(
#  estimate_households_mean_income_dollars < 48500 ~ "Low",
#      estimate_households_mean_income_dollars %in% 48500:145500 ~ "Middle",
#      estimate_households_mean_income_dollars > 145500 ~ "High"
#    )) 

# mutate(
#    income_class =
#     case_when(
#  estimate_households_median_income_dollars < 48500 ~ "Low",
#      estimate_households_median_income_dollars %in% 48500:145500 ~ "Middle",
#      estimate_households_median_income_dollars > 145500 ~ "High"
#    )) 

top_disparity = 
  income_df %>% 
  select(county, state, disparity_value) %>% 
  arrange(desc(disparity_value)) %>% 
  top_n(10) %>% 
  knitr::kable()

disparity_plot =
  income_df %>% 
    ggplot(aes(x = disparity_value)) + 
    geom_histogram()

disparity_plot2 =
  income_df %>% 
  group_by(state) %>% 
  summarise(
    mean_disparity = mean(disparity_value)
  ) %>% 
  top_n(-10) %>% 
  mutate(
    state = fct_reorder(state, mean_disparity)
  ) %>% 
  ggplot(aes(x = state, y = mean_disparity, fill = state)) + 
  geom_col() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

disparity_plot3 =
  income_df %>% 
  group_by(state) %>% 
  summarise(
    mean_disparity = mean(disparity_value)
  ) %>% 
  top_n(10) %>% 
  mutate(
    state = fct_reorder(state, mean_disparity)
  ) %>% 
  ggplot(aes(x = state, y = mean_disparity, fill = state)) + 
  geom_col() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

mpv_add = 
mpv_final %>% 
  mutate(
    county = paste(county, "County", sep = " "),
    county = case_when(
      state == "LA" ~ str_replace(county, "County", "Parish"),
      state == "PR" ~ str_replace(county, "County", "Municipio"),
      TRUE ~ as.character(county)
  ))

```

#### ACS Race

```{r}
race_df =
  read_csv("./data/acs_race_county.csv", skip = 1, na = "null") %>% 
  janitor::clean_names() %>% 
  separate(geographic_area_name, 
           c("county", "state"), sep = ",") %>% 
  select(-c(starts_with("margin_of_error"),
            estimate_total_two_or_more_races,
            estimate_total_two_or_more_races_two_races_including_some_other_race,
            estimate_total_two_or_more_races_two_races_excluding_some_other_race_and_three_or_more_races)) %>% 
  rename(
    total_pop = estimate_total,
    white = estimate_total_white_alone,
    black = estimate_total_black_or_african_american_alone,
    am_in_alask = estimate_total_american_indian_and_alaska_native_alone,
    asian = estimate_total_asian_alone,
    hawaii_pi = estimate_total_native_hawaiian_and_other_pacific_islander_alone,
    other = estimate_total_some_other_race_alone
  ) %>% 
  mutate(
    prop_white = white / total_pop * 100,
    prop_black = black / total_pop * 100,
    prop_am_in_alask = am_in_alask / total_pop * 100,
    prop_asian = asian / total_pop * 100,
    prop_hawaii_pi = hawaii_pi / total_pop * 100,
    prop_other = other / total_pop * 100) %>% 
  select(
    id, total_pop, white, prop_white, black, prop_black, asian, prop_asian, am_in_alask, prop_am_in_alask, hawaii_pi,
    prop_hawaii_pi, other, prop_other
  )
```






