---
title: "Plotting for MPV"
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggplot2)
library(readr)
library(readxl)
library(patchwork)
#you'll need to install the packages below in order to run animated plots or knit
library(gganimate)
library(gifski)
library(png)

knitr::opts_chunk$set(
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_color_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

## MPV Dataset

```{r, warning=FALSE}
mpv_df =
  read_csv("./data/mpv_final.csv")
```

The plot below visualizes the total number of people killed by police, arranged by police department.

```{r}
police_dept_plot = 
  mpv_df %>% 
  count(police_dept) %>% 
  rename(total_killed_pd = n) %>% 
  filter(total_killed_pd >= 49) %>% 
  mutate(
    police_dept = factor(police_dept),
    police_dept = fct_reorder(police_dept, total_killed_pd)
  ) %>% 
  ggplot(aes(x = police_dept, y = total_killed_pd, fill = police_dept)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

police_dept_plot
```

The plots below visualize police killings by state. One includes all states with data, the other includes the top 10 states. 

```{r}
state_plot = 
  mpv_df %>% 
  count(state) %>% 
  rename(total_killed_st = n) %>% 
  arrange(desc(total_killed_st)) %>% 
  mutate(
    state = factor(state),
    state = fct_reorder(state, total_killed_st)
  ) %>% 
  ggplot(aes(x = state, y = total_killed_st, fill = state)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

state_subset_plot = 
  mpv_df %>% 
  count(state) %>% 
  rename(total_killed_st = n) %>% 
  arrange(desc(total_killed_st)) %>% 
  filter(total_killed_st >= 230) %>% 
  mutate(
    state = factor(state),
    state = fct_reorder(state, total_killed_st)
  ) %>% 
  ggplot(aes(x = state, y = total_killed_st, fill = state)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

state_plot + state_subset_plot
```

The plot below shows police killings over time until 2019 (I excluded 2020, because otherwise it looks like 2020 has a drop in police killings).

```{r}
year_plot = 
  mpv_df %>% 
  count(year) %>% 
  rename(total_killed = n) %>% 
  filter(year < 2020) %>% 
  ggplot(aes(x = year, y = total_killed)) + geom_point() +
  geom_smooth(method = lm, se = FALSE) +
  scale_y_continuous(limit = c(0, 1400))

year_plot
```

Code below shows the breakdown of charges laid against police.

```{r}
mpv_case = 
  mpv_df %>% 
  count(criminal_charges) %>% 
  mutate(
    charges = case_when(
      startsWith(criminal_charges, "Charged") ~ "Yes",
      startsWith(criminal_charges, "No") ~ "No",
    )) %>% 
  ggplot(aes(x = charges, y = n)) +
  geom_col() 
```

## Animated Plots

1. Scatter plot showing killings over time separated by race

```{r}
anim_scatter =
  mpv_df %>% 
  group_by(year, race) %>% 
  count() %>% 
  ggplot(aes(x = year, y = n, color = race)) +
  geom_line() +
  geom_point() +
  labs(title = "Number of Police Killings Over Time by Race",
       x = "Year",
       y = "Number of Police Killings") +
  transition_reveal(year)

animate(anim_scatter, renderer = gifski_renderer())

# might be worth excluding 2020 because of incomplete data, it makes it look like the shootings are trending down
```

2. Bar chart showing number of killings by state over time

```{r}
#don't run this!! It's messed up and bogs down computer

bar_format =
  mpv_df %>%
  group_by(year, gender) %>%
  count() %>% 
  mutate(rank = rank(-n),
         Value_rel = n/n[rank==1],
         Value_lbl = paste0(" ",n)) %>%
  group_by(gender) %>%
  filter(rank <= 10)


bar_anim =
  ggplot(bar_format, aes(rank, group = gender))+
  geom_tile(aes(y = n/2,
                height = n,
                width = 0.9), alpha = 0.8, color = NA) +
  geom_text(aes(y = 0, label = paste(gender, " ")), vjust = 0.2, hjust = 1, size = 7) +
  geom_text(aes(y= n,label = Value_lbl, hjust=0),size = 8 ) + 
  coord_flip(clip = "off", expand = TRUE) +
  scale_x_reverse() +
  theme_minimal() +
  theme(axis.line=element_blank(),
        axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        legend.position="none",
        panel.background=element_blank(),
        panel.border=element_blank(),
        panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
        panel.grid.major.x = element_line( size=.1, color="grey" ),
        panel.grid.minor.x = element_line( size=.1, color="grey" ),
        plot.title=element_text(size=25, hjust=0.5, face="bold",     colour="red", vjust=-1),
        plot.subtitle=element_text(size=18, hjust=0.5, face="italic", color="red"),
        plot.caption =element_text(size=12, hjust=0.5, face="italic", color="red"),
        plot.background=element_blank(),
        plot.margin = margin(1,4, 1, 8, "cm")) +
  transition_states(states = year, transition_length = 4, state_length = 1) +
  ease_aes('sine-in-out') +
  labs(title = 'Number of Individuals Killed by Police between 2013 and 2020 in the US: {closest_state}')

final_animation = animate(bar_anim, 100, fps = 20,  width = 750, height = 950, 
        renderer = gifski_renderer())

final_animation
```

