---
title: "Project Report"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
---

```{r, include=FALSE}
library(tidyverse)
library(ggplot2)
library(readr)
library(readxl)
library(patchwork)
library(leaflet)
library(rvest)
library(plotly)
#you'll need to install the packages below in order to run animated plots or knit
library(gganimate)
library(gifski)
library(png)

knitr::opts_chunk$set(
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_color_viridis_d
scale_fill_discrete = scale_fill_viridis_d

mpv_df =
  read_csv("./data/mpv_final.csv")

protest_data =
  read_csv("./data/usa_tidy.csv")

race_data =
  read_csv("./data/race_tidy.csv")

income_df_tidy =
  read_csv("./data/income_tidy.csv")

income_df =
  read_csv("./data/income2.csv")

mpv_demo_df =
  read_csv("./data/mpv_demo_df.csv")
```

<style type="text/css">
</style>

## Motivation

This project is overall motivated by the killing of George Floyd in 2020 and the protests and demonstrations that followed. Police violence and violence against protesters has been a highly talked about issue in 2020 thus far and we were interested in finding and analyzing data related to police violence and the Black Lives Matter protests.

&nbsp;

### *Motivating Sources*

* "A crisis that began with an image of police violence keeps providing more" from NY Times. [Link](https://www.nytimes.com/2020/06/05/us/police-violence-george-floyd.html)
* "Protests about police brutality are met with wave of police brutality in US" from NY Times. [Link](https://www.theguardian.com/us-news/2020/jun/06/police-violence-protests-us-george-floyd)
* "How George Floyd's death ignited a racial reckoning that shows no sign of slowing down" from CNN. [Link](https://www.cnn.com/2020/08/09/us/george-floyd-protests-different-why/index.html)

&nbsp;

## Questions

The questions for this project are based on the 3 datasets we used that are listed in the next section:

1. What is the spread of police violence over time and space, accounting for basic demographics?
2. What were common characteristics of instances of police violence against individuals, including court rulings of the police involved?
3. What is the geographic spread of Black Lives Matter demonstrations in the US during 2020?
4. How does the spread of Black Lives Matter Protests relate to average income across states and cities in the US?

&nbsp;

## Data
### *Data Sources*
* Mapping Police Violence in the US -- [Link](https://mappingpoliceviolence.org/)
* ACLED US Crisis Monitor -- [Link](https://acleddata.com/special-projects/us-crisis-monitor/)
* American Community Survey -- [Link](https://www.census.gov/programs-surveys/acs/data.html)

&nbsp;

#### *Variables of Interest*
##### **Police Violence Dataset**
* `age`
* `gender`
* `race`
* `police_dept`: the police department responsible for the individual's death
* `year`
* `city`
* `state`
* `criminal_charges`: whether the police officers involved received criminal charges
* `symptoms_of_mental_illness`: whether the individual killed had symptoms of mental illness


##### **Protest Dataset**
* `month`
* `event_type`: the type of event, ranging from protests to riots
* `sub_event_type`: the sub-type of event, including peaceful protests
* `state`
* `county`
* `fatalities`: number of individuals killed at each event


##### ** American Community Survey (ACS) Datasets**  

**General**

* `id`
* `county`
* `state`

**Income**

* `total_income`
* `estimate_households`: percent estimate for each county by total household income
* `mean_income`: the mean income in each county
* `median_income`: the median income in each county
* `disparity_value`: mean income - median income in each county, which was constructed in this analysis
* `income_class`: ranges from "Very Low" to "Very High" based on `median_income` value for each county, recoded in this analysis

**Race**

* `race` : classified as white, Black, Asian, Native American, Pacific Islander, and other
* `total`: total estimate for each county for each race category
* `prop`: percent estimate for each county for each race category

&nbsp;

### *Data Cleaning*
The police violence and protest datasets were both similarly cleaned. The data was first imported into R. The variables were then renamed and selected based on if they were intuitive and if they were needed for this analysis. While the protest data already had a latitude and longitude included, an external source was joined with the police violence dataset so latitudes and longitudes of cities could be included in the dataset.

For the ACS data we looked at race and income; each category had their own separate datasets with a tidy and non-tidy version. Tidy datasets were used for visualization and exploratory analyses and the non-tidy datasets were joined with the Police Violence Dataset and Protest Dataset separately in order to conduct further statistical analyses and plotting. However, for both the tidy and non-tidy datasets: variable names were shortened for convenience and only a subset of of the original datasets' categories were kept (for example, in all datasets 'margin of error estimates' were removed). Below are the specific data cleaning steps for each dataset:

* **Race:** This dataset excludes data for those who identify as more than one race. Additionally, proportion was calculated for each race category.
* **Income:** This dataset includes only the total percent estimates by household income. Variables for income class and disparity value were created from the mean and median income of each county.

&nbsp;

## Exploratory Analysis

### *ACS Data*
<br>The exploratory analysis for ACS data was used to provide preliminary context for current country demographics and supplement our analysis of police violence and protests throughout the country.

&nbsp;

#### **US Population by Race**
<br> The two plots below show the total population of each racial group in the US in total (top) and by state (bottom). This breakdown contextualizes the police violence data. Comparing the police killings by race to the total population by race shows us that proportionally, black people have a higher likelihood to be a victim of police brutality.

```{r, warning = FALSE}
mpv_demo_df %>% 
  select(-c(17:18, 21:35, 38, 40, 42, 44, 46, 48)) %>% 
  pivot_longer(
    20:25,
    names_to = "total_race",
    values_to = "total_estimate"
  ) %>%
  mutate(
    total_race = str_replace(total_race, "am_in_alask", "Native American"),
    total_race = str_replace(total_race, "hawaii_pi", "Pacific Islander"),
    total_race = str_replace(total_race, "asian", "Asian"),
    total_race = str_replace(total_race, "black", "Black"),
    total_race = str_replace(total_race, "other", "Other"),
    total_race = str_replace(total_race, "white", "White"),
  )  %>% 
  ggplot(
    aes(x = total_race, y = total_estimate, color = total_race, fill = total_race)
  ) +
  geom_bar(stat = "identity") +
  theme(
      axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 1),
      legend.position = "none") +
  labs(
    x = "",
    y = "Total",
    Title = "Total US Population Estimates by Race"
  )
```

&nbsp;

#### **Household Income in the US**
<br> These plots show the mean household income in the US by state (top) and distribution of disparity in income throughout the US (bottom). As mentioned, the disparity value was taken from subtracting median income from mean income. This visualization supplements our understanding of the distribution and disparity of wealth across the US and provides some background for our regression analyses.

```{r, warning=FALSE}  
# mean household income by state
income_df_tidy %>% 
  mutate(
    state = fct_reorder(state, mean_income)
  ) %>% 
  plot_ly(
      y = ~mean_income, x = ~state, color = ~state,
      type = "box", colors = "viridis")
```

&nbsp;

### *Police Violence Data*
<br>The exploratory analysis for this dataset was primarily used to visualize the data and help in understanding the shape and geographic spread of the data.

&nbsp;

#### **Symptoms of mental illness charges against police (2013-2020)**
<br>The following map shows the geographic distribution of individuals killed by police between 2013 and 2020 by year, with information on symptoms of mental illness and whether the officer was charged included in the label. This map was used to determine the geographic spread of those killed by police each year from 2013 to 2020, and shows the small number of officers that were charged, regardless of killing individuals with symptoms of mental illness.

```{r}
pal <- colorNumeric(
  palette = "viridis",
  domain = mpv_df$year)


mpv_df %>% 
  drop_na() %>% 
  mutate(
    lab = str_c("Symptoms of Mental Illness: ", symptoms_of_mental_illness,
                "<br>Charges on the Police: ", criminal_charges)) %>%
  leaflet() %>% 
  addTiles() %>% 
  setView(-98.483330, 38.712046, zoom = 4) %>% 
  addLegend("bottomright", pal = pal,
            values = ~year,
            title = "Year",
            labFormat = labelFormat(big.mark = ""),
            opacity = 1) %>% 
  addCircleMarkers(
    ~lng, ~lat,
    color = ~pal(year),
    radius = 0.5,
    popup = ~lab)
```

This informed the regression analysis we performed using the police violence data in the regression section below, along with the plot outlined below.

&nbsp;

#### **People killed due to police violence by race**
<br>The following plot displays the number of individuals killed by police over time by race. According to the data displayed in the plot below, the most killings have occurred among white people, followed by black and Hispanic individuals, according to record. This is inconsistent with many findings, which could point to inadequate record of individuals killed by police. This plot along with the map above were used in informing what the regression analysis would examine for this project.

```{r}
anim_scatter =
  mpv_df %>% 
  mutate(
    race = str_replace_all(race, "unknown race", "Unknown race")
    ) %>% 
  filter(year < 2020) %>% 
  group_by(year, race) %>% 
  count() %>% 
  ggplot(aes(x = year, y = n, color = race)) +
  geom_line() +
  geom_point() +
  labs(title = "Number of Police Killings Over Time by Race",
       x = "Year",
       y = "Number of Police Killings") +
  transition_reveal(year)

animate(anim_scatter, renderer = gifski_renderer())
```

&nbsp;

### *Protest Data*

#### **Demonstrations by State**
<br>According to the plot below, California by far held the most demonstrations, followed by New York state and Florida. This informed the regression analysis we performed using the protest data, and gave us the idea to examine this data along with ACS income data on the county level across the US.

```{r}
protest_data %>% 
  mutate(
    state = as.factor(state)
  ) %>% 
  filter(event_type == "Protests") %>% 
  count(state) %>% 
  mutate(
    state = fct_reorder(state, n, .desc = FALSE)
  ) %>% 
   plot_ly(
      x = ~state,
      y = ~n, 
      color = ~state,
      type = "bar", colors = "viridis") %>% 
    layout(
    xaxis = list(title = "State"),
    yaxis = list(title = "Number of Protests"))
```

&nbsp;

## Regression Analysis

### *Police Violence Data*

We opted to investigate two separate outcomes, using this data. The primary model explored whether race was associated with charges laid against police as a result of police violence. Logistic regression was used to investigate this association, as the outcome was binary. We included confounders in the model based on both *a priori* hypotheses and a >10% change in the estimate. Using these criteria, we adjusted for gender and recorded symptoms of mental illness. We then conducted a sensitivity analysis, excluding pending investigations which would have acted to produce misclassification of the outcome. Once we omitted pending cases, a similar association was observed between race and charges laid against police.

The second model assessed the relationship between gender and reported symptoms of mental illness as an outcome. Using the same criteria as outlined above for confounder selection. After adjustment for age and race, we found no significant association between gender and reported symptoms of mental illness.

### *Protest Data*

We investigated protest information at the county-level, using count data which grouped the number of protests by county. Using the dataset linked with census data, we explored whether income class at the county level was associated with number of protests. Since we are interested in modeling count data, we employed Poisson regression with a log link function and no offset term. This model adjusted for population size at the county-level. Upon investigating model fit using overdispersion testing, we found that overdispersion was a concern for this model. Therefore, we ultimately fit a negative binomial model quantifying the association between income class and protest count, adjusted for county population size.

&nbsp;

## Discussion

Overall, the findings for the police violence data were not all consistent with what we had originally hypothesized. Descriptively, we found that the vast majority (98.3%) of police violence cases resulted in no criminal charges, which we had expected. Surprisingly, among people who had recorded symptoms of mental illness, only 0.8% of cases resulted in criminal charges. 

While we did find that most of the cases did not result in criminal charges to the officer involved in the exploratory analysis, the analysis investigating the relationship between race and criminal charges revealed that officers involved in the killing of Black people were more likely to be charged. This could have been for reasons outside of what the data offered, such as these incidents being more frequently unjustified in cases involving Black individuals in comparison to white people. We also found that white individuals had higher numbers of deaths recorded than any other race. However, we were unable to determine whether this is a complication of reporting and classification of cause of death; it is feasible that misclassification as a result of cause of death and recording police deaths may be related to race. Ascertainment of misclassification was not possible with the data available. 

The second model built investigated whether there exists a relationship between gender and recorded symptoms of mental illness. After adjustment for age and race, we did not find a relationship between gender and mental illness.

&nbsp;

There were similarly unexpected but interesting findings in the protest data analysis as well. In the exploratory analysis we found that states with larger populations and large cities, including California, New York, and Florida, held the highest numbers of demonstrations which was consistent with what was expected. However, the Poisson regression analysis found that the "very high" income group was more likely to hold protests than other income groups. We thought that this could have potentially been due to the cities holding more protests in conjunction with having higher income classes, which would be consistent with the finding from the exploratory analysis. After adjustment for population size at the county-level in the original model, we found an association between higher income class and likelihood of holding protests. 

However, after investigating model fit, we found that overdispersion was a concern in the original Poisson regression model. We then fit a negative binomial model to account for overdispersion, and adjusted for county-level population. Once doing so, we found that the only income class significantly associated with a greater protest count was the "low" income class in comparison to the "very low" income class.

Overall, we believe that our results speak to the impact of available resources (in terms of both time and )

we had expected that lower income classes would be more likely to hold demonstrations, which did not hold true here. We do not have the exact reasons for the protests that are in the data so that could potentially be helpful in determining why the higher income was initially recorded as holding significantly more protests.

&nbsp;

**Limitations**

Both datasets contained limited information, which prevented us from being able to adjust for variables we believed could be confounding the associations. In the case of the police violence dataset, the association between race and charges laid was the opposite from what we were expecting. We hypothesized that this could be due to motivation/reason for the initial police stop being less justified when the victim of police violence is Black. Ideally, we would have been able to analyze this relationship by reason for initial police contact. Additionally, we had small cell sizes for people of races other than white or Black, and thus restricted to individuals reporting race as white or Black. In terms of the model assessing income class and protest count, we were limited by the number of covariates available to include. 

&nbsp;

## Conclusions

Overall, our findings were not completely consistent with what we had expected. Our expectations were drawn from what is discussed by the government and by the news, as that is information that is circulated widely. However, there were many inconsistencies we found through our analyses. We also found that given the limitations in our datasets, we may have been missing information that could have benefited our analyses. Police violence data in the past has been somewhat inconsistent and therefore not always trusted, given potential for lack of record of individuals killed by police. However, given the events of 2020, we feel that overall the data was for the most part as complete as the organizations that provided it could have made it.
