---
title: "Data Wrangling for All Datasets Used"
output: 
 html_document:
    code_folding: hide
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
library(tidyverse)
library(readxl)
library(readr)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_color_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

## MPV Dataset

```{r, warning=FALSE}
mpv_df =
  read_excel("./data/MPVDatasetDownload.xlsx",
             sheet = "2013-2020 Police Killings") %>%
  janitor::clean_names() %>% 
  separate(date_of_incident_month_day_year, c("year", "month", "day")) %>% 
  mutate(
    age = as.numeric(victims_age),
    gender  = victims_gender,
    race = victims_race,
    police_dept = agency_responsible_for_death,
    description = a_brief_description_of_the_circumstances_surrounding_the_death,
    disposition = official_disposition_of_death_justified_or_other,
    year = as.numeric(year),
    month = as.numeric(month),
    day = as.numeric(day)) %>%
  select(
    age, gender, race, police_dept, disposition, year,
    month, day, city, state, county, cause_of_death, criminal_charges,
    symptoms_of_mental_illness) %>% 
  view()
```

In the above code chunk, the MPV dataset has been manipulated and tidied by reading in the excel sheet and cleaning the variable names, separating out the date variable, changing the names of some of the variables and changing some variables to be numeric, and selecting only the variables of importance to our analysis for inclusion in the tidied dataset.

## Lat/Long Dataset

```{r}
lat_long =
  read_csv("./data/uscities.csv") %>% 
  mutate(state = state_id) %>% 
  select(city, lat, lng, state)

mpv_final = left_join(mpv_df, lat_long, by = c("city" = "city", "state" = "state"))

write_csv(mpv_final, "./data/mpv_final.csv")
```

## USA October 2020 Dataset

```{r usa, warning=FALSE}
usa_df =
  read_excel("./data/usa_2020_oct.xlsx") %>%
  janitor::clean_names() %>% 
  separate(event_date, c("year", "month", "day")) %>%
  mutate(
    state = admin1,
    county = admin2,
    city = location,
    latitude = as.numeric(latitude),
    longitude = as.numeric(longitude),
    year = as.numeric(year),
    month = as.numeric(month),
    day = as.numeric(day),
    fatalities = as.numeric(fatalities)) %>%
  select(
    year, month, day, event_type, sub_event_type, state, county, city, latitude, longitude, fatalities)

write_csv(usa_df, "./data/usa_tidy.csv")
```

## ACS Datasets

#### ACS Income

Tidy income dataset:
```{r}
income_df =
  read_csv("./data/acs_income_county.csv", skip = 1) %>% 
  janitor::clean_names() %>% 
  separate(geographic_area_name, c("county", "state"), sep = ",") %>% 
  select(
      -c( 
         starts_with("margin_of_error"), 
         starts_with("estimate_families"), 
         starts_with("estimate_married"), 
         starts_with("estimate_nonfamily"), 
         starts_with("estimate_households_percent"))
      ) %>%
  mutate(
    disparity_value = estimate_households_mean_income_dollars - estimate_households_median_income_dollars,
    median_income = estimate_households_mean_income_dollars,
    mean_income = estimate_households_median_income_dollars
    ) %>% 
  pivot_longer(
    5:14,
    names_to = "total_income",
    names_prefix = "estimate_households_total_",
    values_to = "estimate_households"
  ) %>% 
  select(-c(estimate_households_total, estimate_households_mean_income_dollars, estimate_households_median_income_dollars)) %>%
  relocate(c(id, county, state, total_income, estimate_households, mean_income, median_income)) %>% 
  mutate(
    total_income = str_replace_all(total_income, "_", " "),
    total_income = str_replace_all(total_income, "less than 10 000", "0-10,000"),
    total_income = str_replace_all(total_income, "200 000 or more", "200,000+"),
    total_income = str_replace_all(total_income, " to ", "-"),
    total_income = str_replace_all(total_income, " ", ","),
    total_income = as.factor(total_income)
  ) %>% 
  mutate(
   income_class =
    case_when(
     median_income < 55000 ~ "Very Low",
     median_income %in% 55000:59999 ~ "Low",
     median_income %in% 60000:64999 ~ "Middle",
     median_income %in% 65000:74999 ~ "High",
     median_income > 75000 ~ "Very High"
   ),
   income_class = fct_relevel(income_class, "Very Low", "Low", "Middle", "High", "Very High")
  ) %>% 
  view()

# export tidy income dataset
write_csv(income_df, "./data/income_tidy.csv")
```

Income dataset (not tidy):
```{r}
income_df =
  read_csv("./data/acs_income_county.csv", skip = 1) %>% 
  janitor::clean_names() %>% 
  separate(geographic_area_name, c("county", "state"), sep = ",") %>% 
  select(
      -c( 
         starts_with("margin_of_error"), 
         starts_with("estimate_families"), 
         starts_with("estimate_married"), 
         starts_with("estimate_nonfamily"), 
         starts_with("estimate_households_percent"))
      ) %>%
  mutate(
    disparity_value = estimate_households_mean_income_dollars - estimate_households_median_income_dollars) %>% 
mutate(
   income_class =
    case_when(
     estimate_households_median_income_dollars < 55000 ~ "Very Low",
     estimate_households_median_income_dollars %in% 55000:59999 ~ "Low",
     estimate_households_median_income_dollars %in% 60000:64999 ~ "Middle",
     estimate_households_median_income_dollars %in% 65000:74999 ~ "High",
     estimate_households_median_income_dollars > 75000 ~ "Very High"
   ),
   income_class = fct_relevel(income_class, "Very Low", "Low", "Middle", "High", "Very High")
  ) %>%
  mutate(
    county = str_replace(county, " County", ""),
    county = str_replace(county, " Parish", ""),
    county = str_replace(county, " Borough", ""),
    county = str_trim(county, side = "right"),
    state = str_trim(state, side = "left"))

lat_long2 =
  read_csv("./data/uscities.csv") %>% 
  mutate(state = state_id) %>% 
  select(city, lat, lng, state, state_name)

mpv_final2 = left_join(mpv_df, lat_long2, by = c("city" = "city", "state" = "state"))

#combines MPV data and non-tidy income data
build_df = 
  left_join(mpv_final2, income_df, 
            by = c("county" = "county", "state_name" = "state"))

#MPV data and non-tidy income data csv
write_csv(build_df, "./data/build_df.csv")

```

#### ACS Race

Tidy race dataset:

```{r}
race_df =
  read_csv("./data/acs_race_county.csv", skip = 1, na = "null") %>% 
  janitor::clean_names() %>% 
  separate(geographic_area_name, 
           c("county", "state"), sep = ",") %>% 
  select(-c(starts_with("margin_of_error"),
            estimate_total_two_or_more_races,
            estimate_total_two_or_more_races_two_races_including_some_other_race,
            estimate_total_two_or_more_races_two_races_excluding_some_other_race_and_three_or_more_races)) %>% 
  rename(
    total_pop = estimate_total,
    white = estimate_total_white_alone,
    black = estimate_total_black_or_african_american_alone,
    am_in_alask = estimate_total_american_indian_and_alaska_native_alone,
    asian = estimate_total_asian_alone,
    hawaii_pi = estimate_total_native_hawaiian_and_other_pacific_islander_alone,
    other = estimate_total_some_other_race_alone
  ) %>% 
  pivot_longer(
    5:10,
    names_to = "race",
    values_to = "total"
  ) %>% 
  mutate(
    prop = total / total_pop * 100
    ) %>% 
  view()

# export tidy race dataset
write_csv(race_df, "./data/race_tidy.csv")
```


Race dataset (not tidy):

```{r}
race_df2 =
  read_csv("./data/acs_race_county.csv", skip = 1, na = "null") %>% 
  janitor::clean_names() %>% 
  separate(geographic_area_name, 
           c("county", "state"), sep = ",") %>% 
  select(-c(starts_with("margin_of_error"),
            estimate_total_two_or_more_races,
            estimate_total_two_or_more_races_two_races_including_some_other_race,
            estimate_total_two_or_more_races_two_races_excluding_some_other_race_and_three_or_more_races)) %>% 
  rename(
    total_pop = estimate_total,
    white = estimate_total_white_alone,
    black = estimate_total_black_or_african_american_alone,
    am_in_alask = estimate_total_american_indian_and_alaska_native_alone,
    asian = estimate_total_asian_alone,
    hawaii_pi = estimate_total_native_hawaiian_and_other_pacific_islander_alone,
    other = estimate_total_some_other_race_alone
  ) %>% 
  mutate(
    prop_white = white / total_pop * 100,
    prop_black = black / total_pop * 100,
    prop_am_in_alask = am_in_alask / total_pop * 100,
    prop_asian = asian / total_pop * 100,
    prop_hawaii_pi = hawaii_pi / total_pop * 100,
    prop_other = other / total_pop * 100) %>% 
  select(
    id, county, state, total_pop, white, prop_white, black, prop_black, asian, prop_asian, am_in_alask, prop_am_in_alask, hawaii_pi,
    prop_hawaii_pi, other, prop_other
  ) %>%
  mutate(
    county = str_replace(county, " County", ""),
    county = str_replace(county, " Parish", ""),
    county = str_replace(county, " Borough", ""),
    county = str_trim(county, side = "right"),
    state = str_trim(state, side = "left"))

write_csv(race_df2, "./data/race2.csv")
```

## Joining (Non-tidy) ACS data with protest data:

```{r}
race_df =
  read_csv("./data/race2.csv")

income_df =
  read_csv("./data/income2.csv") %>% 
  select(-estimate_households_total) %>% 
  rename_at(
    vars(starts_with("estimate_households_total")), funs(str_replace(., "estimate_households_total_", "")),
    ) %>% 
  rename_at(
        vars(starts_with("estimate_households")), funs(str_replace(., "estimate_households_", ""))
  ) %>% 
  view()

usa_protest_df =
  read_csv("./data/usa_tidy.csv")

# joining protest data with acs data (nontidy)
protest_demo_df = 
  left_join(usa_protest_df, income_df,
            by = c("county" = "county", "state" = "state")) %>% 
  view()

# export as csv
write_csv(protest_demo_df, "./data/protest_demo_df.csv")
```

Retrieving non-tidy ACS income data and total population for regression analysis:
```{r}
protest_demo_df2 =
read_csv("./data/protest_demo_df.csv") %>% view()

protest_demo_df2 =
  protest_demo_df2 %>%
  select(
    -c(13:22, 27:98, 100:111)
  ) %>% 
  mutate(
    county_pop = total_pop
  ) %>% 
  select(
    -total_pop
  ) %>% 
  relocate(
    1:12, county_pop
  ) %>% view()

write_csv(protest_demo_df2, "./data/protest_demo_df2.csv")
```

## Joining MPV data and non-tidy ACS data
```{r}
race_df =
  read_csv("./data/race2.csv")

mpv_df3 = 
  read_csv("./data/mpv_final.csv")

lat_long3 =
  read_csv("./data/uscities.csv") %>% 
  mutate(state = state_id) %>% 
  select(city, lat, lng, state, state_name)

mpv_final3 = left_join(mpv_df3, lat_long3, by = c("city" = "city", "state" = "state"))

#combines MPV data and non-tidy acs data
mpv_demo_df = 
  left_join(mpv_final3, income_df,
            by = c("county" = "county", "state_name" = "state")) %>% 
  left_join(race_df,
            by = c("county" = "county", "state_name" = "state", "id" = "id")) %>% 
  view()

#MPV data and non-tidy acs data csv
write_csv(mpv_demo_df, "./data/mpv_demo_df.csv")
```


